---
title: Multiple Linear Regression Analysis of NBA Player Salaries
date: "`r Sys.Date()`"
output:
 bookdown::pdf_document2:
   extra_dependencies: ["float"]
---

```{r setup, message=FALSE, results="hide", include=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = TRUE, results="hide", message=FALSE,
                      cache=TRUE, fig.pos = "H", out.extra = "")
if(!require("GGally")) install.packages("GGally")
if(!require("kableExtra")) install.packages("kableExtra")
if(!require("cowplot")) install.packages("cowplot")
library("knitr")
# if(!require("magick")) install.packages("magick")
library("forcats")
library("bookdown")
library("readr")
library("dplyr")
library("patchwork")
```

\newpage

# Contributions

Yizhe Zhao

- Rmd code used to analyze data and generate the plots, as well as the formatting for outputting the final report pdf from Rmd
- Preliminary model results section of the report

Shi Wang

- Formatting, rewriting, citations, and general proofreading of the report

Valentyn Lytvyniuk

- Proofreading, organizing in-person meeting, co-writing introduction and data description, collaborating on the data set and response/predictor variable selection

Cyre Beroncal

- Wrote initial drafts of introduction and data description sections, proposed research question
- Collaborated in selecting and interpreting initial variables used in relation to regression outputs

Jack Zhang

- Assisted in selecting and evaluating potential datasets for analysis
- Wrote the notes in the Group Teamwork Agreement Doc
- Proofreading to ensure clarity and accuracy

# Introduction

An athlete’s earnings in professional sports are influenced by factors beyond their raw performance.
Physicality, popularity, and basketball IQ meet in the NBA. A player’s salary can vary based on what they do on the court, how they promote themselves, the position they play, and the terms of their contract.
This project aims to understand the relationship between NBA player salaries and a player’s pure, on-court contributions, using statistical techniques to identify the main factors behind any changes in player pay.

While the common assumption is that performance drives earnings, modern professional sports organizations now operate at the intersection of entertainment and profit. 
This means that, along with more obvious statistics like scoring, playmaking, and defense, team role, fame, and how much a player is worth are also taken into account. 
As a result, our central research question arises as such: "To what extent can NBA player salary be explained by on-court performance and position? What is the relationship between salary, and statistics such as points per game, fouls per game, assists per game, steals per game, and position?"

Linear regression is an appropriate statistical tool for this analysis because we are interested in estimating the average change in salary associated with unit changes in predictor variables (points per game, assists). The method allows us to quantify the effect of each variable while holding others constant. Additionally, categorical predictors such as a player's main position can be incorporated, enabling the model to account for role-based differences in compensation.

By interpreting the linear regression model parameters, it would allow us to understand both player valuation (in terms of their performance) and the patterns that shape the NBA labor market. Could it be, for instance, that certain positions on the court inherently compensated greater, regardless of how successful a player is? Does a player’s earnings better correlate with points or assists than with defensive stats? Such inquiries are important for teams sorting through their salary allocations, players negotiating salaries and those analyzing the fairness of pay scales.

# Data Description

## Data Source

The dataset analyzed here was constructed by Davide Ratto (Ratto, 2019), combining statistics from several public sources available through a project undertaken at a Milan-based university. Data from websites such as basketball-reference.com (Basketball Statistic, n.d.), hoopshype.com (HoopsHype, n.d.), and a Kaggle repository (Gift, 2017) were combined to build this dataset. The objective of the initial study was to create forecasts for player selection to the All-Star Game using a combination of data on performance and popularity. Only the most recent season’s data was kept for each player to preserve a sample that operators of the datasets consider independent.

In the model, we use Salary (USD) as the response variable to investigate which aspects of player performance and role most correlate to NBA compensation. Salary is a continuous and non-negative variable.

## Variables

We're analyzing the response variable, players' Salary (Figure \@ref(fig:vars).A) with these predictors:

- Points Per Game (PTS): Measures a player's scoring ability and offensive productivity (Figure \@ref(fig:vars).B)
- Personal Fouls Per Game (PF): Reflective of a player’s defensive discipline and quality (Figure \@ref(fig:vars).C)
- Assists Per Game (AST): Reflects offensive role and playmaking contributions (Figure \@ref(fig:vars).D)
- Steals Per Game (STL): Indicator of defensive activity and quality (Figure \@ref(fig:vars).E)
- Main Position (Pos1: [PG, SG, SF, PF, C]): A categorical variable that reflects role on court, adding context to the above continuous predictors.

```{r clean}
nba_final <- read.csv("nba_final.csv")
nba_final <- nba_final |>
  filter(Salary>0) |>
  mutate(Salary = Salary / 1e6) # change unit to salary in millions
write.csv(nba_final, file = "nba_final_cleaned.csv")
```

```{r vars, fig.cap='Density histograms of response variables and predictors'}
n <- nrow(nba_final)
nba.hist.Salary <- ggplot(nba_final) +
  geom_histogram(aes(Salary, y = after_stat(density)), 
                 bins = diff(range(nba_final$Salary) / (2 * IQR(nba_final$Salary) / n ^ (1/3)))
                 |> round()) +
  geom_density(aes(Salary)) +
  labs(title = "Salary\nDensity Histogram") +
  labs(x = "Salary\n(USD Millions)") +
  theme(plot.title = element_text(hjust = 0.5, size = 11))

nba.hist.PTS <- ggplot(nba_final) +
  geom_histogram(aes(PTS, y = after_stat(density)), 
                 bins = diff(range(nba_final$PTS) / (2 * IQR(nba_final$PTS) / n ^ (1/3)))
                 |> round()) +
  geom_density(aes(PTS)) +
  labs(title = "Points Per Game\nDensity Histogram") +
  labs(x = "Points per game (PTS)") +
  theme(plot.title = element_text(hjust = 0.5, size = 11))

nba.hist.PF <- ggplot(nba_final) +
  geom_histogram(aes(PF, y = after_stat(density)),
                 bins = diff(range(nba_final$PF) / (2 * IQR(nba_final$PF) / n ^ (1/3)))
                 |> round()) +
  geom_density(aes(PF)) +
  labs(title = "Personal Fouls per game\nDensity Histogram") +
  labs(x = "Personal Fouls (PF)") +
  theme(plot.title = element_text(hjust = 0.5, size = 11))

nba.hist.AST <- ggplot(nba_final) +
  geom_histogram(aes(AST, y = after_stat(density)),
                 bins = diff(range(nba_final$AST) / (2 * IQR(nba_final$AST) / n ^ (1/3)))
                 |> round()) +
  geom_density(aes(AST)) +
  labs(title = "Assists per game\nDensity Histogram") +
  labs(x = "Assists (AST)") +
  theme(plot.title = element_text(hjust = 0.5, size = 11))

nba.hist.STL <- ggplot(nba_final) +
  geom_histogram(aes(STL, y = after_stat(density)), 
                 bins = diff(range(nba_final$STL) / (2 * IQR(nba_final$STL) / n ^ (1/3)))
                 |> round()) +
  geom_density(aes(STL)) +
  labs(title = "Steals per game\nDensity Histogram") +
  labs(x = "Steals per game (STL)") +
  theme(plot.title = element_text(hjust = 0.5, size = 11))

plot_grid(nba.hist.Salary, nba.hist.PTS, nba.hist.PF, nba.hist.AST, nba.hist.STL, labels = "AUTO")

```

## Variable Summaries

```{r scatter-matrix, fig.cap="Scatterplot matrix of respoonse and continuous predictors", fig.height=3.5}
ggpairs(nba_final,
        columns = c("PTS", "PF", "AST", "STL", "Salary"))
```

We have filtered out players with missing salary information. Salary unit is also re-scaled to be in millions.

Final row of Figure \@ref(fig:scatter-matrix) shows that the response variable (Salary) is roughly linear in the other 4 predictor variables. Figure \@ref(fig:box-plots) box plots of Salary against each categorical indicators has different Q2, Q3, and IQR values, except for power forward (PF) and shooting guard (SG). We theorize that once we merge PF and SG into the same level, the model (Equation \@ref(eq:lm)) would be a good linear approximation.
\begin{equation}
Salary = \beta_0 + \beta_1\ PTS + \beta_2\ PF + \beta_3\ AST + \beta_4\ STL + \beta_5\ I(C)+ \beta_6\ I(PG)+ \beta_7\ I(SF) + \varepsilon
(\#eq:lm)
\end{equation}


```{r box-plots, fig.cap="Box Plots of Salary against categorical dummy variables", fig.height=3}
ggplot(nba_final) +
  geom_boxplot(aes(x = Salary, y = Pos1)) +
  labs(x = "Salary (USD Millions)", y = "Player Position (Pos1)")
```

# Preliminary Results

## Residual Analysis

```{r lin-model}
nba_final <- nba_final |>
  # change the team name into categorical levels
  mutate(Pos1 = factor(Pos1)) |>
  # group SG and PF into a single level
  mutate(Pos1 = fct_collapse(Pos1, "SG+PF" = c("SG", "PF"))) |>
  # use shooting guard as reference level
  mutate(Pos1 = relevel(Pos1, ref = "SG+PF"))
nba_model <- lm(Salary ~ PTS + PF + AST + STL + Pos1, data=nba_final)
```

```{r diagnostic-1, fig.cap = "Diagnostic Plots"}
summary(nba_model)
graph_summaries_1 <- function(){
  par(mfrow=c(1, 2))
  p1 <- plot(nba_model, which=1)
  p2 <- plot(nba_model, which=2)
}
graph_summaries_1()
```

```{r diagnostic-2, fig.cap = "Diagnostic Plots, continued", fig.height=4}
graph_summaries_2 <- function(){
  p3 <- ggplot(nba_model) +
    geom_point(aes(x = .fitted, y = nba_final$Salary)) +
    labs(x = "Fitted Values") +
    labs(y = "Response \n Salary") +
    geom_abline(color = "blue") +
    labs(title = "Response vs. Fitted") +
    theme(plot.title = element_text(hjust = 0.5))
  p4 <- ggplot(nba_model) +
    geom_histogram(aes(x = rstandard(nba_model)), bins = 20) +
    theme(plot.title = element_text(hjust = 0.5)) +
    labs(title = "Standarized Residuals Histogram") +
    labs(x = "Standardized Residuals") +
    labs(y = "Count")
  p3 + p4
}
graph_summaries_2()
```

We notice a few signs that our model violates linear regression assumptions. The residual vs. fitted graph (Figure \@ref(fig:diagnostic-1), left) has a pattern, which indicates that the residuals (and therefore errors) are correlated. It also contains points with uneven spread, which indicates heteroscedasticity (i.e. non-constant variance of errors). From the same graph, since the expected value of the residuals is not 0 across the range, non-linearity of our model is also apparent, i.e. Equation \@ref(eq:lm) does not hold.

The model also violates the *normality of error* assumption. The normal-QQ plot (Figure \@ref(fig:diagnostic-1), right) contains points above the identity line from +1 standard deviation and above. The standardized residuals histogram (Figure \@ref(fig:diagnostic-2), right) also shows a long right-tail (right-skewed). This means our model tends to underestimate player salaries at higher salary ranges.

The scatter plot of response vs. fitted values (Figure \@ref(fig:diagnostic-2), left) also suggest that our model is not reliable for predicting salary in terms of the predictors we defined.

## Model Discussion

```{r prelim-model, fig.cap = "Prelimiary estimation of model parameters", results='markup'}

apd <- sprintf("$\\\\hat{\\\\beta_%d}$", seq(from = 0, to = 7))
est_table <- bind_rows(coef(nba_model))
kable(est_table, format = "latex", booktabs = TRUE, escape = FALSE, caption = "Prelimiary estimation of model parameters.") |>
  kable_styling(latex_options = "hold_position") |>
  add_header_above(apd, escape = FALSE)
```

Because of the violations mentioned above, the interpretations we can make are limited. With that in mind, these still estimates tell us that player performance is correlated with their salary, which is also affected by their on-court positions.

In particular, we notice from the estimated model parameters (as seen in Table \@ref(tab:prelim-model)) that the predictor, assists (AST), contribute to the average salary by about \$1.1 million ($\hat{\beta_3} \approx 1.09$) for each unit increase while holding all other predictors constant. Personal fouls (PF) has an opposite effect ($\hat{\beta_2} \approx -0.90$).

For the categorical variables, we expect the average player salary to increase by about \$2.8 million ($\hat{\beta_5} \approx 2.83$) if they change position from shooting guard or power forward (the reference level, SG+PF) to Centre (C), holding all other predictors constant.

\newpage

# Bibliography

*Basketball Statistics & History of every Team & NBA and WNBA players*. Basketball-Reference. (n.d.). [https://www.basketball-reference.com/](https://www.basketball-reference.com/)

Gift, N. (2017). *Social Power NBA*. Kaggle. [https://www.kaggle.com/datasets/noahgift/social-power-nba](https://www.kaggle.com/datasets/noahgift/social-power-nba)

HoopsHype, & Choch, C. A. de. (n.d.). *Hoopshype*. HoopsHype. [https://hoopshype.com/](https://hoopshype.com/)

Ratto, D., Galli, A., & Rubini, R. (2019). *NBA players 2016-2019* [Data set]. Kaggle. [https://www.kaggle.com/datasets/davra98/nba-players-20162019](https://www.kaggle.com/datasets/davra98/nba-players-20162019)
