---
title: Multiple Linear Regression Analysis of NBA Player Salaries
date: "`r Sys.Date()`"
output:
 bookdown::pdf_document2: default
---

```{r setup, message=FALSE, results="hide", include=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = TRUE, results="hide", message=FALSE, cache=TRUE)
if(!require("GGally")) install.packages("GGally")
if(!require("kableExtra")) install.packages("kableExtra")
library("knitr")
# if(!require("magick")) install.packages("magick")
library("forcats")
library("bookdown")
library("readr")
library("dplyr")
library("patchwork")
```

\newpage

# Contributions

# Introduction

# Data Description

## Variables

We're analyzing the players' Salary (response variable) with these predictors:

- Points per game (PTS)
- Personal Fouls per game (PF)
- Assists per game (AST) 
- Steals per game (STL)
- Main position (Pos1), a categorical predictor with 5 levels (SG, C, PF, PG, SF)

## Cleaning data

```{r clean}
nba_final <- read.csv("data/nba_final.csv")
nba_final <- nba_final |>
  filter(Salary>3e5)
write.csv(nba_final, file = "data/nba_final_cleaned.csv")
```

We're filtering out players with missing salary information or wages close to minimum wage for an NBA player, since these are essentially fixed points not affected by any predictors.

```{r scatter-matrix, fig.cap="Scatterplot matrix of respoonse and continuous predictors."}
nba_final <- nba_final |>
  mutate(Salary = Salary / 1e6) # change unit to salary in millions
ggpairs(nba_final,
        columns = c("PTS", "PF", "AST", "STL", "Salary"))
```

Final row of Figure \@ref(fig:scatter-matrix) shows that the response variable (Salary) is roughly linear in the other 4 predictor variables. This indicates that linear regression is suitable here. Figure \@ref(fig:box-plots) shows that the box plots of Salary against each dummy categorical variable (i.e. level) has different Q2, Q3, and IQR values, except for point forward (PF) and shooting guard (SG). This indicates that player position (Pos1) will be a good categorical predictor for Salary after we group PF and SG as one single level.

```{r box-plots, fig.cap="Box Plots of Salary against categorical dummy variables."}
ggplot(nba_final) +
  geom_boxplot(aes(x = Salary, y = Pos1)) +
  labs(x = "Salary (Millions)", y = "Player Position (Pos1)")
```

## Fitting a linear model

```{r factor-collapse}
nba_final <- nba_final |>
  # change the team name into categorical levels
  mutate(Pos1 = factor(Pos1)) |>
  # group SG and PF into a single level
  mutate(Pos1 = fct_collapse(Pos1, "SG+PF" = c("SG", "PF"))) |>
  # use shooting guard as reference level
  mutate(Pos1 = relevel(Pos1, ref = "SG+PF"))
```

```{r prelim-model, fig.cap = "Prelimiary estimation of model parameters.", results='markup'}
nba_model <- lm(Salary ~ PTS + PF + AST + STL + Pos1, data=nba_final)

apd <- sprintf("$\\\\hat{\\\\beta_%d}$", seq(from = 0, to = 7))
est_table <- bind_rows(coef(nba_model))
kable(est_table, format = "latex", booktabs = TRUE, escape = FALSE) |>
  add_header_above(apd, escape = FALSE)
```

```{r diagnostic_1, echo=FALSE, fig.cap = "Diagnostic Plots"}
summary(nba_model)
graph_summaries_1 <- function(){
  par(mfrow=c(1, 2))
  p1 <- plot(nba_model, which=1)
  p2 <- plot(nba_model, which=2)
}
graph_summaries_1()
```

\newpage

```{r diagnostic_2, echo=FALSE, fig.cap = "Diagnostic Plots, continued", fig.height=4}
graph_summaries_2 <- function(){
  p3 <- ggplot(nba_model) +
    geom_point(aes(x = .fitted, y = nba_final$Salary)) +
    labs(x = "Fitted Values") +
    labs(y = "Response \n Salary") +
    geom_abline(color = "blue") +
    labs(title = "Response vs. Fitted") +
    theme(plot.title = element_text(hjust = 0.5))
  p4 <- ggplot(nba_model) +
    geom_histogram(aes(x = rstandard(nba_model)), bins = 20) +
    theme(plot.title = element_text(hjust = 0.5)) +
    labs(title = "Standarized Residuals Histogram") +
    labs(x = "Standardized Residuals") +
    labs(y = "Count")
  p3 + p4
}
graph_summaries_2()
```

## Preliminary Results

### Residual Analysis

We notice a few signs that our model violates linear regression assumptions. The residual vs. fitted graph (\@ref(fig:diagnostic_2) , left) shows a pattern of heteroscedasticity. This makes sense in the context of our data, since players who perform poorly are still guaranteed a minimum salary.
We also notice that the expected value of the residuals is not 0 across the range, which suggests that the residual and fitted values are still correlated in some way, and that our fitted model might not be a linear function of the predictors.

The model also violates the *normality of error* assumption, as shown from the normal-QQ plot (Figure 1, right) and standardized residuals histogram (Figure 2, right). The standardized residuals appear right-skewed, where they have much higher right tail values from the range of +1 standard deviation and above.

The scatter plot of response vs. fitted values (Figure 2, left) show that the response variable does not seem to be any recognizable function of the predictors. This combined with the apparent heteroscedastic pattern in Figure 1 (left) suggests that our model is mis-specified, likely due to inaccurate salary data or other extant factors that the dataset does not account for.

### Model Discussion

Because of the violations mentioned above, the interpretations we make are limited. With that in mind, we notice from the estimated model parameters that the predictor, assists (AST), contribute to the average salary by about \$1 million for each unit increase while holding all other predictors constant. Personal fouls (PF) has an opposite effect. 

On the other hand, for categorical variables, we expect the average player salary to increase by about \$4 million if they change position from shooting guard (SG) to Centre (C), holding all else constant, or decrease by about \$2 million from SG to point guard (PG).
